---
title: "Penguin Report"
author: "Jack Westmoreland (200394930)"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(palmerpenguins)
library(gridExtra)
library(dplyr)
library(knitr)
library(moments)
library(broom)
library(tidyr)
library(kableExtra)

data("penguins")
penguins <- na.omit(penguins) # Removes missing rows
```

# 1. Exploritory Data Analysis
Before conducting any statistical analysis we must first conduct some exploratory
data analysis on our data set. Firstly, i will take a look at the fields the
data set contains.

```{r viewcolumns, echo=FALSE}
colnames(penguins)
```
The data set consists of measurements taken from penguins from the Anvers Islands.
eight measurements of each penguin were taken:

- **species**: The species the penguin belongs to
- **island**: The island the penguin was measured on
- **bill_length_mm**: The length of the penguin's bill (millimeters)
- **bill_depth_mm**: The depth of the penguin's bill (millimeters)
- **flipper_length_mm**: The length of the penguin's flipper (millimeters)
- **body_mass_mm**: The body mass penguin (grams)
- **sex**: The sex of the penguin (male or female)
- **year**: The year the penguin was added to the data set

```{r getsample, echo=FALSE}
n <- 200
my.student.number <- 200394930 # Replace this with your student number
set.seed(my.student.number)
my.penguins <- penguins[sample(nrow(penguins), n), ]
```

### 1.1 Summary Statistics
Using R we can gather some summary statistics on the penguins in the sample
data set.
```{r summarystatistics, echo=FALSE}
summary(my.penguins)
```
From this we can see a few things. Firstly, most penguins in the sample \(P\) set are
from the species Adele and Gentoo, with fewer Chinstrap penguins being found. 
Most of these penguins were found on the islands Biscoe and Dream. Perhaps there
is some correlation between species and island. The sample set contains a roughly
even split of male and female penguins (being only slightly male skewed) and 
the penguins were recorded in the data set between 2007-2009. 

We can also see the sample mean \(\bar{x}\) of the quantitative attributes taken
from the penguins. At a cursory glance we can also see that for these values the
maximum and minimum values in the sample set do not lie massively far away. 
This hints at the data set not containing any outliers, however to further assess
this more accurate more accurate measures must be taken. Which will be done
next by plotting our sample set in histograms.

### 1.2 Penguin Measurment Plots
```{r allpenguinsplot, fig.width=6.5, fig.height=1.6, fig.align='center', fig.cap="All Penguin Measurements", echo=FALSE}
p1 <- ggplot(my.penguins, aes(y = bill_length_mm)) +
  geom_boxplot() +
  labs(title = "Bill Length", y = "Bill Length (mm)", x = "") +
  theme_minimal() + 
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank())

p2 <- ggplot(my.penguins, aes(y = bill_depth_mm)) +
  geom_boxplot() +
  labs(title = "Bill Depth", y = "Bill Depth (mm)", x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank())

p3 <- ggplot(my.penguins, aes(y = flipper_length_mm)) +
  geom_boxplot() +
  labs(title = "Flipper Length", y = "Flipper Length (mm)", x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank())

p4 <- ggplot(my.penguins, aes(y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Body Mass", y = "Body Mass (g)", x = "") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  # Remove x-axis text
        axis.ticks.x = element_blank())

grid.arrange(p1, p2, p3, p4, nrow = 1)
```

From plotting the four quantitative measurements taken from \(P\) we can
see that our measurements are all fairly symmetrical. This is apparent due to \(Q_1\)
and \(Q_3\) of our data being a similar distance from our sample's median.
Furthermore, we can see that the sample does not exhibit any outliers, as there 
are no values outside the ranges:
\[
\left[ Q_1 - 1.5 \times IQR, \, Q_3 + 1.5 \times IQR \right]
\]
We can also see that  bill length, flipper length and body mass all have a fairly 
large interquartile range indicating there is some spread in the data set. This 
can investigated by checking our sample standard deviation (\(s\)) for our variables.

```{r standarddeviationcheck, echo=FALSE}
bill_length_sd <- sd(my.penguins$bill_length_mm, na.rm = TRUE)
bill_depth_sd <- sd(my.penguins$bill_depth_mm, na.rm = TRUE)
flipper_length_sd <- sd(my.penguins$flipper_length_mm, na.rm = TRUE)
body_mass_sd <- sd(my.penguins$body_mass_g, na.rm = TRUE)

cat("Standard Deviation of Bill Length (mm):", bill_length_sd, "\n")
cat("Standard Deviation of Bill Depth (mm):", bill_depth_sd, "\n")
cat("Standard Deviation of Flipper Length (mm):", flipper_length_sd, "\n")
cat("Standard Deviation of Body Mass (g):", body_mass_sd, "\n")
```
As we can see there is some sample standard deviation in our bill/flipper length 
and body mass variables. This is most noticeable for body mass where our samples 
are on average 825 grams from the body mass mean (\( \bar{x}\)). \(s\) is also 
fairly large for bill length and flipper length, whereas bill depth is 
concentrated at it's \(\bar{x}\)

### 1.3 Difference in Species
We could expect that different species of penguins could have different mean \(\bar{x}\)
measurements for our sample data set attributes. To check this we can plot box plots
comparing between the species.
```{r plotbyspecies, fig.width=8, fig.height=2, fig.align='center', fig.cap="Penguin Measurements by Species", fig.pos="H", echo=FALSE}
p1 <- ggplot(my.penguins, aes(x = species, y = bill_length_mm)) +
  geom_boxplot() +
  labs(title = "Bill Length", x = "Species", y = "Bill Length (mm)") +
  theme(axis.text.x = element_text(size = 6))


p2 <- ggplot(my.penguins, aes(x = species, y = bill_depth_mm)) +
  geom_boxplot() +
  labs(title = "Bill Depth", x = "Species", y = "Bill Depth (mm)") +
  theme(axis.text.x = element_text(size = 6))

p3 <- ggplot(my.penguins, aes(x = species, y = flipper_length_mm)) +
  geom_boxplot() +
  labs(title = "Flipper Length", x = "Species", y = "Flipper Length (mm)") +
  theme(axis.text.x = element_text(size = 6))

p4 <- ggplot(my.penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Body Mass", x = "Species", y = "Body Mass (g)") +
  theme(axis.text.x = element_text(size = 6))

grid.arrange(p1, p2, p3, p4, nrow = 1)

```
As we can see there is large variation of these measurements in our sample set \(P\)
depending on the species. For example Gentoo penguins seems to have much
longer flippers, weigh more, but have a smaller bill depth than the other penguins.
This makes penguin species an important measurement when attempting to calculate
our expected values of these measurements for these penguins.

# 2. Fitting A Distribution

From sample set \(P\) it would be helpful to fit a probability distribution for
some of the penguins' measurements. This will allow for us to estimate the
probability of penguins outside the sample holding specific measurements. To
do this I will first plot a histogram of body mass for our sample set \(s\), which
should give a rough idea of what distribution describes the sample data.

```{r histplotallspecies, fig.width=3.5, fig.height=2, fig.align='center', fig.cap="Body Mass Histogram for all Species", echo=FALSE}
ggplot(my.penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200, fill = "skyblue", color = "black") +
  labs(title = "Body Mass Distribution for Penguins", 
       x = "Body Mass (g)", 
       y = "Frequency") +
  theme_minimal()
```
From plotting body mass we can see that the distribution of the sample data
does not seem to resemble any continuous distributions. This is likely due to the
large variation of body mass between species we observed in Figure 2.
Therefore splitting our penguin species up and plotting the distribution by species
may get us better results.

From the sample set, only 45/200 of our penguins are Chinstrap species. Therefore
to give a more robust statistic for this species it would be beneficial to combine
them with the Adelie species. This would increase the sample size for statistical
analysis. To check if this is viable, a t-test can be performed to check for
statistical differences between these species' body mass. First we must
conduct a Barlett test to check if the variance for these sets are equal.

```{r BarlettTestSpecies, echo=FALSE}
adelie <- subset(my.penguins, species == "Adelie")
chinstrap <- subset(my.penguins, species == "Chinstrap")

combined_data <- data.frame(body_mass_g = c(adelie$body_mass_g, chinstrap$body_mass_g),
                            species = factor(rep(c("Adelie", "Chinstrap"), 
                            c(nrow(adelie), nrow(chinstrap)))))
bartlett.test(body_mass_g ~ species, data = combined_data)

```
Our Bartlett gives us a p-value of 0.15, which is significantly large. Therefore
we can assume that the null hypothesis \(H_0\), that these is no significant variance
\(\sigma^2\) differences between the species body masses, holds and we can proceed with a
t-test for body mass between adelie and chinstrap species.

```{r t-testSpecies, echo=FALSE}
results <- t.test(adelie$body_mass_g, chinstrap$body_mass_g, var.equal = TRUE)
p_value <- results$p.value
conf_interval <- results$conf.int

cat("P-value:", p_value, "\n")
cat("95% Confidence Interval:", conf_interval, "\n")
```
From this two sample t-test we can see that the p value is significantly above
the 0.05 threshold for saying there is significant differences in these species
body size. Additionally, the 95% confidence interval of this test contains 0.
Further indicating there are no significant differences. This should allow for
the combination of these species when estimating their body mass probability
distributions.

### 2.1 Normal Distribution.
```{r histplotbyspecie, fig.width=8, fig.height=2, fig.align='center', fig.cap="Body Mass Histogram for Gentoo and combined Adelie,Chinstrap Penguins",  echo=FALSE}
p1 <- ggplot(my.penguins[my.penguins$species == "Gentoo", ], aes(x = body_mass_g)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  labs(title = "Body Mass Distribution for Gentoo Penguins", 
       x = "Body Mass (g)", 
       y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 9, hjust = 0.5))

p2 <- ggplot(my.penguins[my.penguins$species %in% c("Adelie", "Chinstrap"), ],
       aes(x = body_mass_g)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  labs(title = "Body Mass Distribution for Adiele and Chinstrap Penguins", 
       x = "Body Mass (g)", 
       y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(size = 9, hjust = 0.5))

grid.arrange(p1, p2, nrow = 1)
```
These histograms seem to be resemble a normal distribution. Therefore it could
prove useful to fit to our data. This can be further  investigated by using 
quantile-quantile plots to visually see if the values from \(G\) and 
(\( C \cup A \)) resemble the normal distributions. If the plotted quantile
closely align with the normal distribution quantile line it is likely that this
is the case.
```{r qqplots, fig.width=8, fig.height=3, fig.align='center', fig.cap="QQPlots for body mass normal distribution" ,echo=FALSE}
gentoo_penguins <- my.penguins %>% filter(species == "Gentoo")
chinstrap_adelie_penguins <- my.penguins %>% filter(species %in% c("Chinstrap", "Adelie"))
par(mfrow = c(1, 2))

qqnorm(gentoo_penguins$body_mass_g, main = "Gentoo QQ Plot of Body Mass", 
       cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8, cex = 0.7)
qqline(gentoo_penguins$body_mass_g, col = "red")

qqnorm(chinstrap_adelie_penguins$body_mass_g, main = "Chinstrap & Adelie QQ Plot of Body Mass",
       cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8, cex = 0.7)
qqline(chinstrap_adelie_penguins$body_mass_g, col = "blue")

par(mfrow = c(1, 1))
```
From our QQplots we can further verify that both our Gentoo set \(G\) and the
combined Adelie Chinstrap set \(C\cup A\) seem to resemble a normal distribution.
To fit this distribution we must first calculate the mean \(\mu\) and variance 
that \(\sigma^2\) best describe the normal distribution of \(G\) and \(C\cup A\).
To do this the log likelihood function or normal distribution can be used,
\[
\log L(\mu, \sigma^2 \mid x) = -\frac{n}{2} \log(2\pi) - \frac{n}{2} \log(\sigma^2) - \frac{1}{2\sigma^2} \sum_{i=1}^{n} (x_i - \mu)^2
\]
The minimized negative log likelihood function for \(G\) and \(C\cup A\) will
be calculated using R.

```{r minimiseloglikelyhood, echo=FALSE}
neg_log_likelihood <- function(params, data) {
  mu <- params[1]
  sigma_sq <- params[2]
  n <- length(data)
  
  # Calculate the negative log-likelihood
  nll <- -(-n/2 * log(2 * pi) - n/2 * log(sigma_sq) - sum((data - mu)^2) / (2 * sigma_sq))
  return(nll)
}

gentoo_data <- gentoo_penguins$body_mass_g

# Initial guesses for mu and sigma^2
init_params <- c(mean(gentoo_data), var(gentoo_data))

# Perform optimization to maximize likelihood
gentoo_fit <- optim(init_params, neg_log_likelihood, data = gentoo_data, method = "L-BFGS-B", lower = c(-Inf, 0))

gentoo_mu <- gentoo_fit$par[1]
gentoo_sigma_sq <- gentoo_fit$par[2]

chinstrap_adelie_data <- chinstrap_adelie_penguins$body_mass_g

init_params2 <- c(mean(chinstrap_adelie_data), var(chinstrap_adelie_data))

chinstrap_adelie_fit <- optim(init_params2, neg_log_likelihood, data = chinstrap_adelie_data, method = "L-BFGS-B", lower = c(-Inf, 0))

chinstrap_adelie_mu <- chinstrap_adelie_fit$par[1]
chinstrap_adelie_sigma_sq <- chinstrap_adelie_fit$par[2]

gentoo_sample_mean <- mean(gentoo_penguins$body_mass_g)
gentoo_sample_variance <- var(gentoo_penguins$body_mass_g)

ChinstrapAdelie_sample_mean <- mean(chinstrap_adelie_penguins$body_mass_g)
ChinstrapAdelie_sample_variance <- var(chinstrap_adelie_penguins$body_mass_g)
```
**\( G \) \(\hat{\mu}\):** `r round(gentoo_mu, 2)`. 
**\( G \) \(\hat{\sigma}^2\):** `r round(gentoo_sigma_sq, 2)`

**\( G \) \(\bar{x}\):** `r round(gentoo_sample_mean, 2)`. 
**\( G \) \(s^2\):** `r round(gentoo_sample_variance, 2)`

**\( (C \cup A) \) \(\hat{\mu}\):** `r round(chinstrap_adelie_mu, 2)`.
**\( (C \cup A) \) \(\hat{\sigma}^2\):** `r round(chinstrap_adelie_sigma_sq, 2)`

**\( (C \cup A) \) \(\bar{x}\):** `r round(ChinstrapAdelie_sample_mean, 2)`. 
**\( (C \cup A) \) \(s^2\):** `r round(ChinstrapAdelie_sample_variance, 2)`

From maximizing the log normal distribution likelihood function we get the
variables for both these sample sets' normal distribution variables. We can check
these against the sample mean \(\bar{x}\) and sample variance \(s^2\) of these
sets to see that they are identical. This is expected as the recommended estimators
for normal distribution are actually these values.

```{r fittingnormaldist, fig.width=8, fig.height=2, fig.align='center', fig.cap="Fitted Normal Distributions", echo=FALSE, warning=FALSE}
x_values_gentoo <- seq(min(gentoo_penguins$body_mass_g), max(gentoo_penguins$body_mass_g), length.out = 100)
fitted_gentoo <- dnorm(x_values_gentoo, mean = gentoo_mu, sd = sqrt(gentoo_sigma_sq))
fitted_gentoo_df <- data.frame(x = x_values_gentoo, y = fitted_gentoo)

p1 <- ggplot(gentoo_penguins, aes(x = body_mass_g)) +
  geom_histogram(aes(y = ..density..), bins = 20, fill = "lightblue", color = "black") +
  geom_line(data = fitted_gentoo_df, aes(x = x, y = y), color = "red", size = 1.2) +
  labs(title = "Fitted Normal Distribution for Gentoo Penguins' Body Mass",
       x = "Body Mass (g)", 
       y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),)


x_values_chinstrap_adelie <- seq(min(chinstrap_adelie_penguins$body_mass_g), max(chinstrap_adelie_penguins$body_mass_g), length.out = 100)
fitted_chinstrap_adelie <- dnorm(x_values_chinstrap_adelie, mean = chinstrap_adelie_mu, sd = sqrt(chinstrap_adelie_sigma_sq))
fitted_chinstrap_adelie_df <- data.frame(x = x_values_chinstrap_adelie, y = fitted_chinstrap_adelie)

p2 <- ggplot(chinstrap_adelie_penguins, aes(x = body_mass_g)) +
  geom_histogram(aes(y = ..density..), bins = 20, fill = "lightblue", color = "black") +
  geom_line(data = fitted_chinstrap_adelie_df, aes(x = x, y = y), color = "red", size = 1.2) +
  labs(title = "Fitted Normal Distribution for \nChinstrap + Adelie Penguins' Body Mass",
       x = "Body Mass (g)", 
       y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 8, hjust = 0.5),)

grid.arrange(p1, p2, nrow = 1)
```

```{r accuracy, echo=FALSE}
gentoo_skewness <- skewness(gentoo_penguins$body_mass_g)
gentoo_kurtosis <- kurtosis(gentoo_penguins$body_mass_g)

chinstrap_adelie_skewness <- skewness(chinstrap_adelie_penguins$body_mass_g)
chinstrap_adelie_kurtosis <- kurtosis(chinstrap_adelie_penguins$body_mass_g)
```
**\( G \) Skewness:** `r round(gentoo_skewness, 2)`.
**(\( C \cup A \)) Skewness:** `r round(chinstrap_adelie_skewness, 2)`

**\( G \) Kurtosis:** `r round(gentoo_kurtosis, 2)`.
**(\( C \cup A \)) Kurtosis:** `r round(chinstrap_adelie_kurtosis, 2)`

From the fitted plots (fig 6) we can visually see that the normal distribution does a
relatively good job fitting to our data. The histogram destiny looks to fairly
resemble the red normal distribution line fitted to the sets. To further 
test these distributions the skewness and kurtosis of the sets have been measured.
From these measures we can see that \(G\) is slightly right skewed compared to
the normal distribution skew of 0. However this is still quite close to 0 thus
reinforcing this models validity. The skewdness of (\( C \cup A \)) is a little
more, implying a right skew to the data set however this is still fairly close to 0.

By measuring the Kurtosis of \(G\) and (\( C \cup A \)) we can see that
the tails of these sets are slightly heavier than that of the normal distribution.
Meaning that there are more extreme outlines in the sample sets than expected in the
normal distribution, which has a kurtosis of 3. Overall from visual and quantitative 
assessment, sets \(G\) and (\( C \cup A \)) appear to be reasonably close to a 
normal distribution. With the small size of our sets \(\approx 200\), likely 
effecting \(G\) and (\( C \cup A \))s accuracy when approximating the penguin
population. The normal distribution likely is the best fit for our data. However,
due to the slight differences in the normal distribution and the sets the sample
measurements may vary from the true population to some degree.


# 3. Sexing
To find out which characteristics of penguins would be best at estimating their
sex we can first plot a box plot of each penguins' measurements by their sex.\

```{r Sexingbotplot, fig.width=8, fig.height=2, fig.align='center', fig.cap="Penguin Measurements by Sex", echo=FALSE}
p1 <- ggplot(my.penguins, aes(x = sex, y = bill_length_mm)) +
  geom_boxplot() +
  labs(title = "Bill Length", x = "Sex", y = "Bill Length (mm)") +
  theme(axis.text.x = element_text(size = 6))


p2 <- ggplot(my.penguins, aes(x = sex, y = bill_depth_mm)) +
  geom_boxplot() +
  labs(title = "Bill Depth", x = "Sex", y = "Bill Depth (mm)") +
  theme(axis.text.x = element_text(size = 6))

p3 <- ggplot(my.penguins, aes(x = sex, y = flipper_length_mm)) +
  geom_boxplot() +
  labs(title = "Flipper Length", x = "Sex", y = "Flipper Length (mm)") +
  theme(axis.text.x = element_text(size = 6))

p4 <- ggplot(my.penguins, aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Body Mass", x = "Sex", y = "Body Mass (g)") +
  theme(axis.text.x = element_text(size = 6))

grid.arrange(p1, p2, p3, p4, nrow = 1)
```
From these plots we can observe some difference in the median \( X \) for these
measurements depending on their sex. Specifically the male penguins seem
to have a much larger body mass median vs female. This hints at there being 
some statistical difference in penguins from our male subset and female subset.

We can represent the hypotheses for further testing as:

- \( H_0 \): No difference in measurements between males and females, \( \mu_{\text{male}} = \mu_{\text{female}} \).

- \( H_1 \): There is a difference, \( \mu_{\text{male}} \neq \mu_{\text{female}} \).

### 3.1 Hypothesis Testing
To confirm if hypothesis \(h_1 \) holds statistical significance a two-sample 
t-test can be performed on \(M \subset P\) \(F \subset P\) where:

\( M = \{p \in P \mid p.sex = "male"\}\) 

\( F = \{p \in P \mid p.sex = "female"\}\)

For running t-tests the sets \(F\) & \(M\) must be **IID**, independent and
identically distributed. It is safe to say that the penguins measurements are not
effected by each other. One penguin \(p\) will not have its flipper length effected
by another and therefore are independent. \(F\) & \(M\) also have measurements
coming from the same distributions within their sets and are therefore identically
distributed.

First me must check if the sample variance between male and female penguins us 
roughly equal for both sets so that \(\sigma^2_M = \sigma^2_F\). To check this a
a Barlett's test can be performed with R.
```{r BarlettTestSex, echo=FALSE}
bodymassSexBresults <- bartlett.test(body_mass_g ~ sex, data = my.penguins)
bodymassSexBp_value <- bodymassSexBresults$p.value

cat("P-value:", bodymassSexBp_value, "\n")
```
With this result we can accept the null hypothesis that there is no difference in
body mass variation for \(F\) and \(M\) as \(pvalue < 0.05\) therefore a two sample
t-test should be conducted.
```{r t-testSex, echo=FALSE}
male <- subset(my.penguins, sex == "male")
female <- subset(my.penguins, sex == "female")

bodymassSexTresults <- t.test(male$body_mass_g, female$body_mass_g, var.equal = TRUE)
bodymassSexTp_value <- bodymassSexTresults$p.value
bodymassSexTconf_interval <- bodymassSexTresults$conf.int
bodymassSexTformatted_p_value <- format(bodymassSexTp_value, scientific = FALSE)


cat("Sex Body Mass Welch Test Result")
cat("P-value:", bodymassSexTformatted_p_value, "\n")
cat("95% Confidence Interval:", bodymassSexTconf_interval, "\n")
```
From this result we can see there is a large difference in body size for our sets
\(F\) and \(M\), with a p value far blow 0.05. The 95% confidence interval puts this
difference around 420-845g. Making this a potentially great marker for distinguishing between
male and female penguins. From performing the same tests on each of the penguins
measurements we get the following.

```{r t-testsAllSex, echo=FALSE}
## Bill Length
BillLenSexBresults <- bartlett.test(bill_length_mm ~ sex, data = my.penguins)
BillLenSexBp_value <- BillLenSexBresults$p.value
BillLenSexTresults <- t.test(male$bill_length_mm, female$bill_length_mm, var.equal = FALSE)
BillLenSexTp_value <- BillLenSexTresults$p.value
BillLenSexTconf_interval <- BillLenSexTresults$conf.int

## Bill Depth
BillDepthSexBresults <- bartlett.test(bill_depth_mm ~ sex, data = my.penguins)
BillDepthSexBp_value <- BillDepthSexBresults$p.value
BillDepthTresults <- t.test(male$bill_depth_mm, female$bill_depth_mm, var.equal = TRUE)
BillDepthTp_value <- BillDepthTresults$p.value
BillDepthTconf_interval <- BillDepthTresults$conf.int

## Flipper Length
FlippLenSexBresults <- bartlett.test(flipper_length_mm ~ sex, data = my.penguins)
FlippLenSexBp_value <- FlippLenSexBresults$p.value
FlippLenTresults <- t.test(male$flipper_length_mm, female$flipper_length_mm, var.equal = TRUE)
FlippLenTp_value <- FlippLenTresults$p.value
FlippLenDepthTconf_interval <- FlippLenTresults$conf.int
```

```{r t-teststable, echo=FALSE}
data <- data.frame(
  Measurement = c("Bill Length", "Bill Depth", "Flipper Length", "Body Mass"),
  `Bartlett's P-value` = c(format(round(BillLenSexBp_value, 4), scientific = FALSE),
                           format(round(BillDepthSexBp_value, 4), scientific = FALSE),
                           format(round(FlippLenSexBp_value, 4), scientific = FALSE),
                           format(round(bodymassSexBp_value, 4), scientific = FALSE)),
  
  `T-test's P-value` = c(format(round(BillLenSexTp_value, 8), scientific = FALSE),
                         format(round(BillDepthTp_value, 8), scientific = FALSE), 
                         format(round(FlippLenTp_value, 8), scientific = FALSE), 
                         format(round(bodymassSexTp_value, 8), scientific = FALSE)),
  
    `95% Confidence Interval (Lower)` = c(BillLenSexTconf_interval[1], 
                                        BillDepthTconf_interval[1], 
                                        FlippLenDepthTconf_interval[1], 
                                        bodymassSexTconf_interval[1]),
  `95% Confidence Interval (Upper)` = c(BillLenSexTconf_interval[2], 
                                        BillDepthTconf_interval[2], 
                                        FlippLenDepthTconf_interval[2], 
                                        bodymassSexTconf_interval[2])
)

kable(data, linestep = "", format = "markdown", col.names = c("Measurement", "Bartlett's P-value", "T-test's P-value", 
                                                "95% Confidence Interval (Lower)", "95% Confidence Interval (Upper)"))
```
This table shows that there are several strong differences in male and female
penguin measurements. Body mass and Bill Depth seem to be the two best measurements
for this. However, the other measurements are all strong candidates, with p-values
far below the 0.05 threshold needed to reject \(H_0\). Furthermore, for all
measurements the 95% confidence intervals places the differences fairly far from 0
meaning these measurements quite reliable for sexing male and female pigeons.
Therefore with a high degree of certainty we can say for all measurements \(H_1\) holds.

### 3.2 Logistic Regression
To actually make predictions of penguin sex we can fit a logistic regression model
to our data using the four measurements. Logistic regression is a statistical model used to 
predict binary outcomes (in this case male or female) from from our variables.
It works by predicting the log odds (the odds of male or female) as a linear function
of our penguins variables. These odds are then turned into a probability of the penguin
belonging to a sex with use of a logistic function. Essentially finding the boundaries
that separate our males and females for these variables. This differs from the two
sample t-tests done in section 3.1 which can only take into account one variable of \(P\)
to determine differences between \(F\) and \(M\)

```{r logistic regression, echo=FALSE}
model <- glm(sex ~ body_mass_g + bill_length_mm + flipper_length_mm + bill_depth_mm, 
             data = my.penguins, 
             family = binomial)
model_summary <- tidy(model)
model_summary <- model_summary %>% select(term, estimate, std.error, statistic, p.value)

predicted_prob <- predict(model, my.penguins, type = "response")
predicted_sex <- ifelse(predicted_prob > 0.5, "male", "female")
accuracy <- mean(predicted_sex == my.penguins$sex)
print(model_summary)
cat("Model Accuracy: ", round(accuracy * 100, 2), "%\n")

```
From fitting the logistic regression model we can that the most significant measures
for determining sex are identified as body mass and bill depth. These values have much lower
p values than others for sex estimation. This is consistent with our two sample analysis.
Furthermore, the \(\sigma\) error and z vales (statistic) are extremely small for these
measures, further proving that there are good variables for predicting sex. The model
achieved from these parameters achieved a 88.5% accuracy, meaning the sex of each
penguin was correctly predicted 88.5% of the time.

# 4. Island Characteristics
The penguin sample set \(P\) has penguins who come from three different islands 
"Biscoe", "Torgersen" and "Dream" to find out if islands have significant statistical
impact on penguin measurements we can split sample set \(P\) into three subsets.
and create our hypothesis.

\( B = \{p \in P \mid p.island = "Biscoe"\}\) 

\(T = \{p \in P \mid p.island = "Torgersen"\}\)

\(D = \{p \in P \mid p.island = "Dream"\}\)

- \( H_0 \): No difference in measurements between \(B\), \(T\) and \(D\), \( \mu_{B} = \mu_{T} = \mu_{D}\).

- \( H_1 \): There is a difference, \( \mu_{B} \neq \mu_{T} \neq \mu_{D}\).

```{r splitIntoIslands, echo=FALSE}
biscoe <- subset(my.penguins, island == "Biscoe")
torgersen <- subset(my.penguins, island == "Torgersen")
dream <- subset(my.penguins, island == "Dream")
```

```{r penguinSpeciesonIsland, fig.width=8, fig.height=3, fig.cap="Species Island Count & Penguin by Island Boxplts", echo=FALSE}
species_count <- as.data.frame(table(my.penguins$island, my.penguins$species))
colnames(species_count) <- c("Island", "Species", "Count")

p1 <- ggplot(species_count, aes(x = Island, y = Count, fill = Species)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Penguin Species Distribution Across Islands", x = "Island", y = "Number of Penguins") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  )

penguins_long <- my.penguins %>%
  pivot_longer(cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
               names_to = "Measurement", values_to = "Value")

p_combined <- ggplot(penguins_long, aes(x = island, y = Value)) +
  geom_boxplot() +
  facet_wrap(~ Measurement, scales = "free_y", ncol = 2) +  # Facet by measurement, 2 columns
  labs(title = "Penguin Measurements by Island", x = "Island", y = "Value") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6),
    strip.text = element_text(size = 10)
  )

grid.arrange(p1, p_combined,nrow = 1)
```
 \(B\), \(T\) and \(D\) all contain a differing amounts of penguins from each species.
 For example Gentoo penguins are only found on \(B\) and Chinstrap on \(D\) in the
 sample set. Therefore, measuring the differences of all penguins on each island (as
 seen in the above box plot) would more than likely instead measure the differences in
 species. Which has already been completed in section 2. Therefore, to ensure I am
 only measuring statistical differences due to the island of the penguins I will be 
 updating the sets to only contain Adelie penguins. Which are present on all 3 islands.
 
\( B = \{p \in P \mid p.island = "Biscoe"\ and \ p.species = "Adelie"\}\) 

\(T = \{p \in P \mid p.island = "Torgersen"\ and \ p.species = "Adelie"\}\)

\(D = \{p \in P \mid p.island = "Dream"\ and \ p.species  = "Adelie"\}\)

```{r adelieboxplotbyIsland, fig.width=8, fig.height=1.8, fig.align='center', fig.cap="Adelie Measurements by Island", echo=FALSE}
p1 <- ggplot(adelie, aes(x = island, y = bill_length_mm)) +
  geom_boxplot() +
  labs(title = "Bill Length", x = "Island", y = "Bill Length (mm)") +
  theme(axis.text.x = element_text(size = 6))


p2 <- ggplot(adelie, aes(x = island, y = bill_depth_mm)) +
  geom_boxplot() +
  labs(title = "Bill Depth", x = "Island", y = "Bill Depth (mm)") +
  theme(axis.text.x = element_text(size = 6))

p3 <- ggplot(adelie, aes(x = island, y = flipper_length_mm)) +
  geom_boxplot() +
  labs(title = "Flipper Length", x = "Island", y = "Flipper Length (mm)") +
  theme(axis.text.x = element_text(size = 6))

p4 <- ggplot(adelie, aes(x = island, y = body_mass_g)) +
  geom_boxplot() +
  labs(title = "Body Mass", x = "Island", y = "Body Mass (g)") +
  theme(axis.text.x = element_text(size = 6))

grid.arrange(p1, p2, p3, p4, nrow = 1)
```
From the box plots of sets \(B\), \(D\) and \(T\) we can see little variation
between the medians and IQR of Adelie penguin measurements between the three islands.

## 4.1 ANOVA Tests
For statistical analysis between the three islands an ANOVA test will be conducted
on each penguin measurement. ANOVA tests work by calculating the between group
variability \(SSB\) and within group variability \(SSW\) for each group (island).

\(SSB = \sum_{i=1}^{k} n_i (\bar{X}_i - \bar{X})^2\)

\(SSW = \sum_{i=1}^{k} \sum_{j=1}^{n_i} (X_{ij} - \bar{X}_i)^2\)

\(k\): number of groups

\(n_i\): number of observations in group i

\(k\): number of groups

\(\bar{X}_i\): mean of group \(i \)

\(\bar{X}\): overall mean

\(X_{ij}\): individual data point \(j\) in group \(i\)

\(SSB\) measures the variation in between groups, by calculating how much the groups
mean differs from the overall mean and multiplying this by the number of observations 
in the group. This is summed for each group (island). \(SSW\) measures the in group variability
by summing the squared difference from the mean for each data point in the group.
This is done and summed for all groups. We then calculate F ratio \(f = \frac{SSB}{SSW}\)
which is used in a look up table to find it associated p value.

ANOVA tests assume that the data is normally distributed, and of equal variance
between groups. From barlett's testing we can see that all data has equal 
variation (p-vale > 0.5 for all measurements). Additionally, by plotting Q-Q plots
like as done in section 2.1 confirms the data is normally distributed. Making
ANOVA tests a good choice for statistical analysis across islands.
```{r ANOVAtest, echo=FALSE}
anova_body_mass <- aov(body_mass_g ~ island, data = adelie)
summary_body_mass <- summary(anova_body_mass)
body_mass_p_value <- summary_body_mass[[1]][["Pr(>F)"]][1]

anova_bill_len <- aov(bill_length_mm ~ island, data = adelie)
summary_bill_len <- summary(anova_bill_len)
bill_len_value <- summary_bill_len[[1]][["Pr(>F)"]][1]

anova_bill_depth <- aov(bill_depth_mm ~ island, data = adelie)
summary_bill_depth <- summary(anova_bill_depth)
bill_depth_value <- summary_bill_depth[[1]][["Pr(>F)"]][1]

anova_flipper_length <- aov(flipper_length_mm ~ island, data = adelie)
summary_flipper_length <- summary(anova_flipper_length)
flipper_length_p_value <- summary_flipper_length[[1]][["Pr(>F)"]][1]

cat("body mass:", round(body_mass_p_value,4), "bill length:", round(bill_len_value, 4)
    ,"bill depth:", round(bill_depth_value, 4), "flipper length:", round(flipper_length_p_value, 4))

```
From the p values, derived from ANOVA tests on each measurement. there is not enough 
statistical difference between adelie penguins on different island to reject null 
hypothesis \(H_0\). The p-values from these tests are too large to suggest the data 
shows significant differences in measurements between islands. The p-value for 
bill depth could indicate that there is variation in measurements between islands, 
but this is still too large to reject \(H_0\). Therefore it is unlikely that there is 
differences in penguin measurements between islands when adjusted for the differences
for species.
